---
title: "Methods (BST 232) Notes"
format: 
  html:
    toc: true
    theme: ../style.scss
---

Welcome to my notes on Methods for Simple, Multiple, and Generalized Linear Models! 

```{r, echo=FALSE, out.width='100%'}
knitr::include_graphics('week2/standalone_figures/normal_equation_projection/normal_equation_projection.svg')
```

I hope you will enjoy them, but you do have to be prepared: the notation is pretty 
all over the place. 

I've been honing my $\LaTeX$ tikz and pgfplots skills throughout this course, 
so I hope you enjoy the diagrams I have diligently been creating throughout. 

Besides the above, here are a couple more figures I'm so thrilled to have been
able to learn how to make in $\LaTeX$.

```{r, echo=FALSE, out.width='100%'}
knitr::include_graphics('week1/standalone_figures/scientific_process/scientific_process.svg')
knitr::include_graphics('week2/standalone_figures/pr_density_of_linear_model/pr_density_of_linear_model.svg')
knitr::include_graphics('week3/standalone_figures/residual_decompositions/residual_decomposition.svg')
```

:::{.cooltip}
You may be asking, why on Earth would you spend your time learning tikz when it's so 
convoluted, tortured, arcane, and generally a pain-in-the-ass?

I have two reasons: 
  
  1. I find it to be nicer for producing mathematical "infographics" (explanatory graphics) than R's `{ggplot2}`
    which I love to death for data visualization. 
  2. [left to the reader as an exercise]
:::